Machine Learning, Building Vintage and Property Values

### 2019-09-04 text/manuscript_assa.tex
Small text changes, Table 1 more condensed

\begin{document}
\maketitle
\begin{abstract}
This paper first introduces an algorithm that collects pictures of
individual buildings from Google Street View. Second, it trains a deep
convolutional neural network (CNN) to classify residential buildings
into architectural styles, taking into account spatial dependencies of
building vintages. Third, it investigates whether architectural styles
influence house prices. For re-sales, the architectural style is a
significant determinant of transaction prices while no such effect is
found for new buildings. This indicates that any premia are for
vintage-related quality characteristics and not for a home's ``beauty''.
\end{abstract}

@@

\footnote{{*Corresponding author: Thies Lindenthal (htl24@cam.ac.uk). Replication files are available from the author's Github repository \href{https://github.com/thies}{https://github.com/thies}. Paul E. Glade and Lukas Heckmann-Umhau are thanked for exceptional research assistance. Carolin Schmidt, Mike Langen, Colin Lizieri, Franz Fuerst, Stanimira Milcheva, D'Maris Coffman, Norm Miller, Stephen Malpezzi, John Clapp and seminar participants at UCL/Bartlett and at the Homer Hoyt Institute offered excellent feedback and critique on earlier versions of this paper.}}

@@

This paper will assess architectural aesthetics heads on and analyze the
visual appearance of residential buildings using street level images.
Recent work illustrates how street level imagery and machine learning
classification is an efficient and powerful combination for measuring
previously unobserved characteristics of the urban environment. Naik et
al. (2017) describe how neighborhood demographics may impact the
physical appearance of neighborhoods. Gebru et al. (2017) use classified
vehicle make and model to predict income, race, education, and voting
patterns at the precinct level. Glaeser et al. (2018) predict income in
New York City. Naik, Raskar, and Hidalgo (2016) create a neighborhood
safety based Streetscore which is shown to be highly correlated with
neighborhood population density and household income. De Nadai et al.
(2016) find that greenery and street facing windows contribute to a
positive appearance of safety while Liu et al. (2017) evaluate the
quality and upkeep of the built environment along Beijing's streets.

@@

The aesthetics of different architectural eras make for an easy cocktail
party conversation topic -- and they also influence market prices. Using
architectural assessments by human experts, Buitelaar and Schilder
(2017) find a sizable premium of 5\% for new buildings in the
Netherlands that refer to traditional styles and a staggering 15\%
premium for new buildings that closely follow traditional shapes, facade
composition and details. Their study carefully disentangles the
architectural style from other unobserved characteristics such as
building quality, differences in location or year of construction. These
controls are crucial, as earlier work has established that age and
vintage variables tend to be highly correlated. Coulson and McMillen
(2008), for instance, suggest a non-parametric estimator and establish a
U-shaped age function and distinct price discounts for postwar and
contemporary vintages (vis-à-vis more historic vintages). Francke and
Minne (2017) investigate the depreciation of residential real estate in
the Netherlands and decompose land versus structure values singling out
the effect of ``physical deterioration, functional obsolescence and and
vintage effects''. They find that buildings from the 1930s carry a
strong price premium.

@@

Unfortunately, the traditional approaches chosen by e.g. Buitelaar and
Schilder (2017) or Ahlfeldt and Mastro (2012) do not scale well. Each
observation has to be classified into architectural styles by a human
expert, which is time-consuming and costly, imposing an upper limit on
the number of observations and the level of detail captured for each
observation in any given sample. Google Street View offers a solution to
this nexus as it captures the images of almost all buildings in many
cities around the world at high level of accuracy and detail. The
challenge we address in this paper is to extract and utilize building
level information from this ubiquitous sensor.

@@

We download all pictures at the highest resolution offered by Google's
Street View
API\footnote{\href{https://developers.google.com/maps/documentation/streetview/intro}{https://developers.google.com/maps/documentation/streetview/intro}}
(640x640 color pixels). We then obtain 2048-dimensional feature vectors
for each picture, using the \emph{Inception-v3} deep convolutional
neural network (Szegedy et al. 2015). \emph{Inception-v3} has been
trained for the ImageNet Large Scale Visual Recognition Challenge
(ILSVRC)\footnote{\url{http://image-net.org/challenges/LSVRC/}}, which
evaluates image classification and object detection algorithms for a
wide range of objects. The pretrained classifications would allow us to
identify pets, vehicles or people on the pictures -- assessing
architectural finesse, however, is beyond the canned classifiers'
capabilities.

@@

Two variables are obtained from the distribution of classifications per
observation: The ensemble classifier arrives at predictions by majority
vote. In addition, a confidence measure is computed as the Herfindahl
index of classifications by ensemble member
models.\footnote{If, for instance, 20 models classified a building as *Contemporary* but 80 others as *Revival*, the ensemble majority vote is *Revival*. The corresponding Herfindahl score is calculated as $(20/100)^2 + (80/100)^2 = 0.68$.}

@@

Residential real estate transactions are public data in the UK,
collected and published by the Land Registry (Land Registry 2017). The
records include the date of transaction, price paid, street address, a
classification of the property type (flat, detached, semi-detached, or
terraced house), the estate type (freehold or leasehold) and an
indicator for newly built properties. We select transactions from
Cambridge proper which were recorded between January 1995 and October
2018, excluding any leaseholds, flats and properties classified as type
``other'', and sales with prices below £50,000 or in excess of
£2,000,000. Linking recently taken images to sales data from 24 years
assumes that houses have not changed their architectural styles -- which
seems reasonable. We are not aware of any conversions of just the
exterior of buildings. Full redevelopments, did occur and we exclude
sales for a given address pre-dating any redevelopment. Table
\ref{tab:sumstats} presents summary statistics for the sample.

@@

The Herfindahl scores for classifications vary substantially across
ground truth vintages (Table \ref{tab:herftab}), indicating that the
models within the ensemble, for instance, agree more often on a building
being \emph{Georgian} than finding consensus for other vintages. For
\emph{Contemporary} or \emph{Revival} vintages, individual
classifications differ most. All off-diagonal elements in the lower
panel of Table \ref{tab:herftab} are negative, suggesting that, in
general, the ensemble is not really sure when misclassifying images.

@@

The number of misclassifications decreases decidedly when considering
observations with high ensemble confidence only: The majority vote based
on the most confident two-thirds of predictions only are more in line
with ground truth, as higher recall, precision and \(F_1\)-scores in
Table \ref{tab:confmat}, Panel B, show. The improvement is most evident
in the precision rates. For the relatively rare categories of
\emph{Georgian} and \emph{Revival} about the same number of buildings
classified as \emph{Georgian/Revival} in Panel A actually belong to a
different ground truth class -- the precision shoots up to 0.70 for
\emph{Georgian} and 0.82 for \emph{Revival} when excluding the not so
confident lower third of predictions in Panel B.

@@

The first column in Table \ref{tab:hedreg} presents the estimated
regression coefficients for ground truth vintages. The base vintage
\emph{Contemporary} is more expensive than almost all other vintages,
which show negative coefficients that are significantly different from
0. A clear pecking order appears: \emph{Georgian} and \emph{Revival}
buildings are the dearest \((+0.04)\), followed by \emph{Late
Victorian/Edwardian} \((+0.01)\), \emph{Contemporary} (base),
\emph{Early Victorian} \((-0.13)\), \emph{Interwar} \((-0.15)\) and
\emph{Postwar} \((-0.22)\).

@@

Introducing a variable capturing the most prominent style of buildings
on the same street and within 100 m (\(\mathbf{VintNeigh}\)) does not
change the pecking order of building vintages much (Table
\ref{tab:hedreg}, Column 7). Neighboring styles are compared to the base
case of a \emph{Contemporary} house in a \emph{Contemporary} street. The
negative coefficients on dissimilar neighboring styles support a finding
by Lindenthal (2017b) that shows that a harmonious match of a building's
shape with its direct environment leads to a price premium -- or a
discount in case of shape mismatches. In relative terms, \emph{Postwar}
buildings are the least popular neighbors. The difference between
\emph{Revival} and \emph{Contemporary} neighbors is not statistically
significant. We fail to find evidence in favor of positive externalities
from revival architecture. Everything else equal, buildings within
ensembles of \emph{Contemporary} buildings generally do not achieve
higher transaction prices than buildings in more historic areas (as
Table \ref{tab:comb} will confirm later).

@@

The contributions of this paper are threefold: First, it introduced an
algorithm that collects pictures of \emph{individual} buildings from
Google Street View. Earlier work has not achieved this level of detail
and was, at least in the UK, limited to street sections only. A
large-scale application of automatic classification of individual
buildings' characteristics using Google Street View has potential not
only in the UK. The image collection and classification method can
easily be ported to other study areas which have existing Street View
data and either LIDAR based building outlines or high resolution
satellite images. In a follow-up project, we are working on an improved
classification workflow in which a customized object detection model is
able to recognize specific building outlines without the need for
additional maps or other data.

>>>>

\begin{document}
\maketitle
\begin{abstract}
This paper first introduces an algorithm that collects pictures of
individual buildings from Google Street View. Second, it trains a deep
convolutional neural network (CNN) to classify residential buildings
into architectural styles, taking into account spatial dependencies of
building vintages. Third, it investigates whether architectural styles
influence house prices. For re-sales, the architectural style is a
significant determinant of transaction prices while no such effect is
found for new buildings. Additionally, through the comparison of expert
classification and deep learning methods, we are able to illustrate some
of the limitations of machine learning methods and provide guidance on
how to detect and overcome these limitations.
\end{abstract}

@@

\footnote{{*Corresponding author: Thies Lindenthal (htl24@cam.ac.uk). Replication files are available from the author's Github repository \href{https://github.com/thies}{https://github.com/thies}. Paul E. Glade and Lukas Heckmann-Umhau are thanked for exceptional research assistance. Carolin Schmidt, Mike Langen, Colin Lizieri, Franz Fuerst, Stanimira Milcheva, D'Maris Coffman, Norm Miller, Stephen Malpezzi, John Clapp, Pavel Krivenko and participants at the ASSA/AREUEA 2019 annual meeting and seminars at UCL/Bartlett and the Homer Hoyt Institute offered excellent feedback and critique on earlier versions of this paper.}}

@@

This paper assesses the value of architectural aesthetics heads on by
classifying the vintage of residential buildings using street level
images. Recent work illustrates how street level imagery and machine
learning classification is an efficient and powerful combination for
measuring previously unobserved characteristics of the urban
environment. Naik et al. (2017) describe how neighborhood demographics
may impact the physical appearance of neighborhoods. Gebru et al. (2017)
use classified vehicle make and model to predict income, race,
education, and voting patterns at the precinct level. Glaeser et al.
(2018) predict income in New York City. Naik, Raskar, and Hidalgo (2016)
create a neighborhood safety based Streetscore which is shown to be
highly correlated with neighborhood population density and household
income. De Nadai et al. (2016) find that greenery and street facing
windows contribute to a positive appearance of safety while Liu et al.
(2017) evaluate the quality and upkeep of the built environment along
Beijing's streets.

@@

The aesthetics of different architectural
vintages\footnote{We use ``style'',``era'' or ``vintage'' interchangeably in this paper.}
make for an easy cocktail party conversation topic -- and they may also
influence market prices. Using architectural assessments by human
experts, Buitelaar and Schilder (2017) find a sizable premium of 5\% for
new buildings in the Netherlands that refer to traditional styles and a
staggering 15\% premium for new buildings that closely follow
traditional shapes, facade composition and details. Their study
carefully disentangles the architectural style from other unobserved
characteristics such as building quality, differences in location or
year of construction. These controls are crucial, as earlier work has
established that age and vintage variables tend to be highly correlated.
Coulson and McMillen (2008), for instance, suggest a non-parametric
estimator and establish a U-shaped age function and distinct price
discounts for postwar and contemporary vintages (vis-à-vis more historic
vintages). Francke and Minne (2017) investigate the depreciation of
residential real estate in the Netherlands and decompose land versus
structure values singling out the effect of ``physical deterioration,
functional obsolescence and and vintage effects''. They find that
buildings from the 1930s carry a strong price premium.

@@

Unfortunately, the traditional approaches chosen by e.g. Buitelaar and
Schilder (2017) or Ahlfeldt and Mastro (2012) do not scale well. Each
observation has to be classified into architectural styles by a human
expert, which is time-consuming and costly, imposing an upper limit on
the number of observations and the level of detail captured for each
observation in any given sample. Google Street View offers a solution to
this nexus as it captures the images of almost all buildings in many
cities around the world at high level of accuracy and detail. The
challenge we address in this paper is to extract and utilize building
level information from this ubiquitous sensor, using freely available
deep learning techniques.

@@

We download all pictures at the highest resolution offered by Google's
Street View
API\footnote{\href{https://developers.google.com/maps/documentation/streetview/intro}{https://developers.google.com/maps/documentation/streetview/intro}}
(640x640 color pixels). We then obtain 2048-dimensional feature vectors
for each picture, using the \emph{Inception-v3} deep convolutional
neural network (Szegedy et al. 2015). \emph{Inception-v3} has been
trained for the ImageNet Large Scale Visual Recognition Challenge
(ILSVRC)\footnote{\url{http://image-net.org/challenges/LSVRC/}}, which
evaluates image classification and object detection algorithms for a
wide range of objects. The pretrained classifications would allow us to
identify pets, vehicles or people on the pictures -- assessing
architectural style, however, is beyond the canned classifiers'
capabilities.

@@

Two variables are obtained from the distribution of classifications per
observation: The ensemble classifier arrives at predictions by majority
vote. In addition, a confidence measure is computed as the Herfindahl
index of classifications by ensemble member models. If, for instance, 20
models classified a building as \emph{Contemporary} but 80 others as
\emph{Revival}, the ensemble majority vote is \emph{Revival}. The
corresponding Herfindahl score is calculated as
\((20/100)^2 + (80/100)^2 = 0.68\).

@@

Residential real estate transactions are public data in the UK,
collected and published by the Land Registry (Land Registry 2017). The
records include the date of transaction, price paid, street address, a
classification of the property type (flat, detached, semi-detached, or
terraced house), the estate type (freehold or leasehold) and an
indicator for newly built properties. We select transactions from
Cambridge proper which were recorded between January 1995 and October
2018, excluding any leaseholds, flats and properties classified as type
``other'', and sales with prices below £50,000 or in excess of
£2,000,000. Linking recently taken images to sales data from 24 years
assumes that houses have not changed their architectural styles -- which
seems reasonable. We are not aware of any conversions of just the
exterior of buildings. Full redevelopments did occur and we exclude
sales for a given address pre-dating any redevelopment. Table
\ref{tab:sumstats} presents summary statistics for the sample.

@@

The Herfindahl scores for classifications vary substantially across
ground truth vintages (Table \ref{tab:herftab}), indicating that the
models within the ensemble, for instance, agree more often on a building
being \emph{Georgian} than finding consensus for other vintages. For
\emph{Contemporary} or \emph{Revival} vintages, individual
classifications differ most. All off-diagonal elements in the lower
panel of Table \ref{tab:herftab} are negative, suggesting that, in
general, the Herfindahl index is a good predictor of misclassified
images.

@@

The number of misclassifications decreases decidedly when considering
observations with high ensemble confidence only: The majority vote based
on the most confident two-thirds of predictions only are more in line
with ground truth, as higher recall, precision and \(F_1\)-scores in
Table \ref{tab:confmat}, Panel B, show. The improvement is most evident
in the precision rates. For the relatively rare categories of
\emph{Georgian} and \emph{Revival} about the same number of buildings
classified as \emph{Georgian/Revival} in Panel A actually belong to a
different ground truth class -- the precision shoots up from 0.50 to
0.70 for \emph{Georgian} and from 0.52 to 0.82 for \emph{Revival} when
excluding the not so confident lower third of predictions in Panel B.

@@

The first column in Table \ref{tab:hedreg} presents the estimated
regression coefficients for ground truth vintages. The base vintage
\emph{Contemporary} is more expensive than almost all other vintages,
which show negative coefficients that are significantly different from
0. A clear pecking order appears: \emph{Georgian} and \emph{Revival}
buildings demand the highest premium \((+0.04)\), followed by \emph{Late
Victorian/Edwardian} \((+0.01)\), \emph{Contemporary} (base),
\emph{Early Victorian} \((-0.13)\), \emph{Interwar} \((-0.15)\) and
\emph{Postwar} \((-0.22)\).

@@

Introducing a variable capturing the most prominent style of buildings
on the same street and within 100 m (\(\mathbf{VintNeigh}\)) does not
change the pecking order of building vintages much (Table
\ref{tab:hedreg}, Column 7). Neighboring styles are compared to the base
case of a \emph{Contemporary} house in a \emph{Contemporary} street. The
negative coefficients on dissimilar neighboring styles support a finding
by Lindenthal (2017b) that shows that a harmonious match of a building's
shape with its direct environment leads to a price premium -- or a
discount in case of shape mismatches. In relative terms, \emph{Postwar}
buildings are the least popular neighbors. The difference between
\emph{Revival} and \emph{Contemporary} neighbors is not statistically
significant. We fail to find evidence in favor of positive externalities
from revival architecture. Everything else equal, buildings within
ensembles of \emph{Contemporary} generally do not achieve higher
transaction prices than buildings in more historic areas (as Table
\ref{tab:comb} will confirm later).

@@

The contributions of this paper are threefold: First, it introduces an
algorithm that collects pictures of \emph{individual} buildings from
Google Street View. Earlier work has not achieved this level of detail
and was, at least in the UK, limited to street sections only. A
large-scale application of automatic classification of individual
buildings' characteristics using Google Street View has potential not
only in the UK. The image collection and classification method can
easily be ported to other study areas which have existing Street View
data and either LIDAR based building outlines or high resolution
satellite images. In a follow-up project, we are working on an improved
classification workflow in which a customized object detection model is
able to recognize specific building outlines without the need for
additional maps or other data.
